Big Ideas in Sports Analytics and Statistical Tools for their Investigation

### 2022-07-18 sports_analytics.Rmd
Some edits.

A more in-depth overview of the history of expected points in sports is provided in @Yurko2019nflwar (Section 1.1). 
Most importantly, this paper uses publicly available data provided by the **nflscrapR** package [@R-nflscrapR] to model the expected points on a play-by-play level in football.
The authors introduce a multinomial logistic regression approach, which takes into account the current down, time remaining, yards from endzone, yards to go, and the indicators for goal to go and whether the time remaining in the half is less than two minutes. 
This model estimates the probabilities of the following possible scoring outcomes after each play: no score, safety, field goal, and touchdown for both the offensive and defensive teams, all of which have a point value. 
The expected points for a play can then be calculated accordingly, by summing up the products of the scoring event point values and their associated probabilities.

### 4th down strategy

@@

4th down strategy papers based on expected points.  
@Yam2019: 
@Romer2006: Examines 4th down decisions in the NFL using expected points.  Focuses only on the first quarter.  Teams don't go for it enough is they are trying to maximize their probability of winning the game.  
@pelechrinis2019 Present an expected points framework where they adjust for the strength of the defense.  
<!-- New york times article: https://www.nytimes.com/2014/09/05/upshot/4th-down-when-to-go-for-it-and-why.html?_r=0 -->

@@

Numerous papers (see @lopez2020bigger for details) have used the analysis of the expected number of points to improve fourth down strategy.

>>>>

A more in-depth overview of the history of expected points in sports is provided in @Yurko2019nflwar (Section 1.1). 
Most importantly, this paper uses publicly available data provided by the **nflscrapR** package [@R-nflscrapR] to model the expected points on a play-by-play level in football.
The authors introduce a multinomial logistic regression approach, which takes into account the current down, time remaining, yards from endzone, yards to go, and the indicators for goal to go and whether the time remaining in the half is less than two minutes. 
This model estimates the probabilities of the following possible scoring outcomes after each play: no score, safety, field goal, and touchdown for both the offensive and defensive teams, all of which have a point value.

The expected points for a play can then be calculated accordingly, by summing up the products of the scoring event point values and their associated probabilities.

### 4th down strategy
The concept of expected points in American football has many applications.  One of the most notable and well-studied is the use of expected points in the evaluation of 4th down strategy, and there is near universal consensus in the literature that NFL teams have been too conservative when making 4th down decisions.  @Romer2006 examines 4th down decisions in the NFL using expected points by focusing only on example from the first quarter of a game (to avoid issues with end of half and end of game decision making).  They conclude that teams don't go for it enough if they are trying to maximize their probability of winning the game.

Numerous papers (see @lopez2020bigger for details) have used the analysis of the expected number of points to improve fourth down strategy. @Romer2006 Examines 4th down decisions in the NFL using expected points.  Focuses only on the first quarter.  Teams don't go for it enough is they are trying to maximize their probability of winning the game. Later work in @Yam2019 used win probability, rather than expected points, and a casual inference framework to reach similar conclusions that NFL teams were to conservative in going for it on 4th down.  In addition, they estimate that a better strategy would have been worth about 0.4 wins per season on average, which is substantial in a season with only 16 games.

Talks about how tracking data is important for a more nuanced discussion of 4th down decisions.

@@

4th down strategy papers based on expected points.  
@lopez2020bigger: Talks about how tracking data is important for a more nuanced discussion of 4th down decisions.  
<!-- @Yam2019: Uses Causal inference techniques to examine the effects of 4th down decision making in the NFL.   -->
<!-- @Romer2006: Examines 4th down decisions in the NFL using expected points.  Focuses only on the first quarter.  Teams don't go for it enough is they are trying to maximize their probability of winning the game.   -->
@pelechrinis2019 Present an expected points framework where they adjust for the strength of the defense.  
<!-- New york times article: https://www.nytimes.com/2014/09/05/upshot/4th-down-when-to-go-for-it-and-why.html?_r=0 -->

@@
